---
title: "Analysis of blight risk for Detroit area"
author: "Paweł Stradowski"
date: "May 2016"
output: html_document
---
```{r loader, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
Sys.setlocale("LC_MESSAGES", "en_US.utf8")
Sys.setlocale("LC_ALL", "en_US.utf8")
#data_path <- '~/datasci_course_materials/capstone/blight/data/'
data_path <- '/home/pawel/Documents/datasci_course_materials/capstone/blight/data/'
detroit.311 <- read.csv(paste0(data_path, "detroit-311.csv"))
violations <- read.csv(paste0(data_path, "detroit-blight-violations.csv"))
crime <- read.csv(paste0(data_path, "detroit-crime.csv"))
demolitions <- read.delim(paste0(data_path, "detroit-demolition-permits.tsv"))
load("boundary.RData")
```

# Introduction

An interactive version of this document is available here <https://rpubs.com/pstradowski/blight>, I strongly recommend to use it, because of the leaflet components, which are non-interactive in the PDF version.  


This if final assignment from Coursera Data Science at Scale specialization.
The aim of this analysis is to build a model, which will predict the risk of blight for certain building or geo coordinates located in Detroit.
The model should take a location and give an answer about likelihood of blight.  
The model takes as train input the events, which could have an impact on possible condemnment:

* crimes
+ violations
+ calls to 311  

This data will be used to train model, the labels for the models will come from data about demolition permits issued by City of Detroit.  
Main part of the analysis is done in R, the data cleaning and enrichement was done using dplyr, visualisation was mostly based on ggplot and leaflet.  
Full source code is available online at <https://github.com/pstradowski/blight> 


```{r libs, echo=FALSE, warning=FALSE, message=FALSE, include=TRUE}
library("dplyr")
library("leaflet")
library("tidyr")
library("magrittr")
library("geohash")
library("aspace")
library("xgboost")
library("ggmap")
library("knitr")
```
```{r enrich, echo=FALSE, warning=FALSE, message=FALSE, include=TRUE}
parse_addr <- function(df, col){
  df %>% 
    separate_(col, into=c("addr", "loc"), sep="\\(") %>%
    tidyr::extract(loc, into = c("lat", "lon"), 
          regex = "(\\d+\\.\\d+),\\s(\\-*\\d+\\.\\d+)", perl = TRUE) %>%
  mutate(lat=as.numeric(lat), lon=as.numeric(lon)) %>%
  mutate(addr = sub("\n.+", "", addr)) 
 }

enrich <- function(df) {
  df %>% 
    mutate(geohash = gh_encode(lat, lon, 9)) %>%
    mutate(gh6 = substr(geohash, 1, 6),
         gh7 = substr(geohash, 1, 7),
         coord = paste0(lat, lon)) 
    
}

filter_outliers <- function(df, epsilon = 4) {
  df %<>% filter(lat<max(boundary$lat), 
                    lat > min(boundary$lat), 
                    lon > min(boundary$lon), 
                    lon < max(boundary$lon))
  cleaned <- df %>% group_by(coord) %>%
    summarise(acnt = n_distinct(addr)) %>%
    filter(acnt <= epsilon) %>%
    select(-acnt)
  df %>% inner_join(cleaned, by = "coord")
}

demolitions %<>% parse_addr("site_location") %>% 
  enrich() 

violations %<>% parse_addr("ViolationAddress") %>% 
  enrich()
viol <-  violations %>% filter_outliers()
  
crime %<>% rename(lat = LAT, lon = LON, addr = ADDRESS) %>% 
  enrich()
crm <-  crime %>% filter_outliers()

detroit.311 %<>% rename(lon = lng) %>%
  mutate(addr = gsub("Detroit.+", "", address)) %>%
  enrich() 
det.311 <- detroit.311 %>%  filter_outliers()
```
This assignment is heavily based on geospatial analysis, therefore I considered following tools to analyze spatial distance:  

* *dplyr with geohash* as quick and dirty solution for joining data based on its spatial attributes
+ *[tile38](https://github.com/tidwall/tile38)* - specialized geolocation datastore which would allow for more precise joins
This analysis covers only geohash usage, in case of further development, I strongly recommend to check also more precise solution based on tile38 or similar datastore.  

# Data Quality

In project materials, there is no information about gathering process of the input data. This makes quality analysis very difficult, because we can find data which is bad, but have no tools to understand the root cause of the problem and correct the problem in a smart way. Instead we will throw out the data which looks troublesome.  
The other issue with the input data is lack of positive labels. We only have records about demolition permits, which mean that we can for sure say that certain building was blighted, but we don't have opposite labels. This makes the whole analysis One Class Classification problem and we can describe it as PU Learning [^1] - we have only Positive Set (demolition) and Unlabeled set (the rest), so it impacts the way, we design and train our model. As next step, I would definitely recommend review of available PU Learning approaches and evaluate them to find best one. For purpose of this document, I will focus on optimalisation for AUC as the main method to deal with One Class Classification problem.  
Data files have different syntax - one needs to transform it to common format for latitude, longitude and addresses. In this step I also calculated geohashes.  
The other issue with the data are outliers, I identified 2 classes of them:

* location outside Detroit
+ multiple addresses assigned to a lat/lon pair - an example of top 5 lat/lon pairs from violatinos database

[^1]: See: <https://en.wikipedia.org/wiki/One-class_classification>
<http://www.machinedlearnings.com/2012/03/pu-learning-and-auc.html>
``` {r, echo=FALSE}
data <- violations %>% group_by(coord) %>%
    summarise(address_count = n_distinct(addr)) %>% arrange(-address_count) %>% dplyr::slice(1:5)
knitr::kable(data, caption = "Top 5 coordinates wit multiple addresses")
```

  
The data was filtered to the area of Detroit using hand selected boundaries and all locations with more than 4 distinct addresses were rejected. This reducded the size of violations by `r nrow(violations)-nrow(viol)` rows and crimes by `r nrow(crime)-nrow(crm)` rows.  

# Visualization and exploration

Spatial distribution of all data is not uniform, there are some high-level clusters.
```{r, echo=FALSE, warning=FALSE, message=FALSE, include=TRUE}
det_map <- ggmap(get_map(location = c(-83.31717, 42.291342, -82.89610, 42.47767), source="osm"))

det_map + scale_fill_continuous(low="orange", high="red") +
  stat_density2d(aes(x = lon, y = lat, fill = ..level..), alpha = 0.2, size = 0.1, data = viol, geom = "polygon") + ggtitle("Density of violations")

det_map + scale_fill_continuous(low="orange", high="red") +
  stat_density2d(aes(x = lon, y = lat, fill = ..level..), alpha = 0.2, size = 0.1, data = crm, geom = "polygon") + ggtitle("Density of crimes")

det_map + 
  scale_fill_continuous(low="orange", high="red") +
  stat_density2d(aes(x = lon, y = lat, fill = ..level..), alpha = 0.2, size = 0.1, data = demolitions, geom = "polygon") + ggtitle("Density of demolition permits") 
```
## Interactive map of demolition permits
```{r, echo=FALSE, warning=FALSE, message=FALSE, include=TRUE}
leaflet(data = demolitions) %>%
  addTiles() %>% 
  addMarkers(~lon, ~lat, clusterOptions = markerClusterOptions()) 

```

```{r, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
blight <-demolitions %>% distinct(geohash) %>%
  select(geohash, gh6, gh7, lat, lon) %>%
  mutate(blighted = 1) 


all_events <- crm %>% select(geohash, gh6, gh7, lat, lon) %>%
  bind_rows(viol %>% select(geohash, gh6, gh7, lat, lon)) %>%
  distinct(geohash) 

non_blight <- anti_join(all_events, blight, by="gh6") %>%
  select(geohash, gh6, gh7, lat, lon) %>%
  mutate(blighted = 0)

data_set <- non_blight %>%
  sample_n(nrow(blight), replace = FALSE) %>% 
  bind_rows(blight)

featurise <- function(df, precision, spread_col){
  df %>% select_("geohash", "gh6", "gh7", "lat", "lon", spread_col) %>%
  inner_join(data_set, by = precision) %>%
  group_by_("geohash.y", spread_col) %>% 
  summarise(cnt = n()) %>%
  rename(geohash = geohash.y) %>%
  spread_(spread_col, "cnt", fill = 0)   
}

glue_it <- function(...){
  tables <- list(...)
  out_table <- data_set
  for (i in tables ) {
    out_table %<>% left_join(i, by = "geohash")
  }
  out_table %<>% replace(is.na(.), 0)
  tr <- out_table %>% 
    select(-lat, -lon, -geohash, -gh7, -gh6, -blighted) %>%
    as.matrix()
lb <- out_table %>%
        select(blighted) %>%
        as.matrix()
xgb.DMatrix(tr, label=lb)
}


feat_viol <- featurise(viol, "gh7", "ViolationCode")
feat_crime <- crm %>% filter(!is.na(STATEOFFENSEFILECLASS)) %>%
  featurise("gh7", "STATEOFFENSEFILECLASS")

xgb_tr <- glue_it(feat_viol, feat_crime)
model_cv <- xgb.cv(data = xgb_tr, objective = "binary:logistic", 
                   eval_metric = "auc",
                     nfold=5, nrounds=50, max_depth=15, eta=0.8,
                     gamma=1, colsample_bytree=0.8,
                     subsample=0.75, min_child_weight=2)
```
# Features
Assignment 4 from the course assumed usage of 1 feature - number of violations associated with the building, in my opinion such feature is very basic and must be extended.  
I belive, that not every crime category and violation category is equal - some of them will be more important for determining blight risk. I split the data by categories - Violation Code and STATEOFFENSEFILECLASS, this gives me one feature for each category of crime and violation.  
The final features will consist of number of crime/violations for each category and location within the same geohash. To deal witth different distances, I used 6 or 7 character geohash to check, which distance gives best results.    
This approach can be made more precise by using specialised software to join datafiles using kind of geospatial indexes i i.e tile38.

# Model
I used Extended Gradient Boost with initial set of parameters used in one of previous projects. In further phase, a hyperparameter could be done to ensure best results.  
Due to PU Learning problem, I decided to use AUC as the evaluation metric, however the PU Learning problem should be further analyzed in order to valiadate assumptions for my approach.  
Test set consists of data from demolition files plus equal amount of locations taken from other sources but with different 6 char geohash. This gives us points which lies more than 600m from blighted homes. This can impact the model performance, so an alternative could be to add an additional feature measuring distance to the next blighted home.

# Results and summary

I made 5 fold cross validation and it finishes with AUC `r model_cv$test.auc.mean[100]` which looks very good.  
In order to ensure that this result is good, one need to repeat the analysis without approximations caused by geohash and PU Learning problem.  
The AUC result strongly depends on selection of unknown points - it is recommended to switch to high-precision selection of points and features instead of geohashes and then add a feature or features  measuring distance to the next blighted house.